---
title: "Lab: Correlation and Multiple Linear Regression with Qualitative Variables"
author: "Zi Ye"
date: "`r Sys.Date()`"
format:
  html:
    toc-depth: 4
---

The lecture’s slides can be found [here](https://github.com/GDSL-UL/stats/blob/main/lectures/lecture03.pdf).

Last week, we introduced Multiple Linear Regression (MLR) - a statistical method that models the relationship between a dependent variable and two or more independent variables, allowing researchers to examine how various predictors jointly influence an outcome. By using the following R code, we create and interpret the model:

`model <- lm(pct_Very_bad_health ~ pct_No_qualifications + pct_Males + pct_Higher_manager_prof, data = census)`

`summary(model)`

In a regression model, independent/predictor variables could be continuous or categorical (or qualitative). While continuous variables capture quantitative effects, **categorical variables provide insights into differences across groups**. When we say categorical variables, we normally mean:

- Nominal Data: categorical data without natural order. E.g. Gender, Colour, Country...

- Ordinal Data: categorical data with a meaningful order. E.g. Education level, Customer satisfaction, Grade...

By blending continuous and categorical predictors, MLR with categorical variables enhances the model's ability to reflect real-world complexities and improves interpretability, as it allows analysts to assess how each category or group within an independent variable influences the dependent variable.

For most categorical (especially the *nominal*) variables, they cannot be included in the regression model directly as a continuous independent variable. Instead, these qualitative independent variables should be included in regression models by using the **dummy variable** approach, transforming categorical information into a numerical format suitable for regression analysis.

However, R provides a powerful way, by automatively handling this process when the categorical variable is designated as a factor and to be included in the regression model. This makes it much easier for you to use categorical variables in the regression model to assess the effects of categorical groupings on the dependent variable alongside continuous predictors.

Learning Objectives:

In this week's practical we are going to

- Analysis of categorical/qualitative variables

- Estimate and interpret a multiple linear regression model with categorical variables

- Make predictions using a regression model

## Analysing categorical variables

Recall in Week 2, you got familiar with R by using the Census data. Today we will explore both the Family Resource Survey (FRS) and the Census data by using their categorical variables. You should already have your Census data in your local drive folder under the path of '/data/Census2021/'; for the FRS datasets, please click the links to download the datasets from [Canvas](https://canvas.liverpool.ac.uk/courses/84668/pages/family-resource-survey-2016-17?module_item_id=2396198). Please download both the datasets of 'FRS16-17_labels.csv' and 'FRS_dictionary.xlxs'<!--and 'FRS16-17_codes.csv'--> for today's practical. You may create a new folder named 'FRS' under the '/data/' along with the Census2021 folder, and put the newly downloaded .csv files at the '/data/FRS/' folder for later use.

<!--Now, there should be two .csv file: 'FRS16-17_labels.csv' and 'FRS16-17_codes.csv' in your '/data/FRS' folder. Double click to open both .csv files you may find they have exactly same column names - yes, they are in fact the same spreadsheet but 'FRS16-17_labels.csv' is more readable for the cell contents as they are in text format but cells 'FRS16-17_codes.csv' is in coding format. This is quite normal as when we doing surveys, we use text as the options of multiple choice questions for respondents to choose from, and therefore the collected survey results are in the 'FRS16-17_labels.csv' format. For example, for the multiple choice question - which region you are from? The options include 'London', 'North West', 'Wales', etc. However, while analysing the survey results, these text content (the options of multiple choice questions) are in fact different categories of the categorical/qualitative variable. So for some survey results, you may find they are formatted in coding way as 'FRS16-17_codes.csv', and along with a coding dictionary for you to learn the meaning of each coding. -->

To start today's practical session, we first will use 'FRS16-17_labels.csv'. <!--In the Region example, there are 12 categories for the categorical/qualitative variable Region--> You can open the .csv file in Excel and find its different from the Census dataset, the FRS dataset contains many qualitative/categorical variables, such as hh_tenure (Housing Tenure), happy (How happy did you feel yesterday?), health (How is your health in general) and etc. To know the meaning of all the column names and the values in cell, you may need to reference to the 'FRS_dictionary.xlsx' to help your understanding. Recall our lecture, these categorical variable will need to be treated as Dummy variable in the regression model. In R, the variable type of qualitative/categorical variables is called '**factor**'.

As usual we first load the necessary libraries.

**Some tips to avoid R returning can't find data errors:**

Check your working directory by

```{r,results='hide',message=FALSE}
getwd()
```

Check the relative path of your data folder on your PC/laptop, make sure you know the relative path of your data from your working directory, returned by `getwd()`.

**Library knowledge used today:**

- **`dplyr`**: a basic library provides a suite of functions for data manipulation

- **`ggplot2`**: a widely-used data visualisation library to help you create nice plots through layered plotting.

- **`tidyverse`**: a collection of R packages designed for data science, offering a cohesive framework for data manipulation, visualization, and analysis. Containing dplyr, ggplot2 and other basic libraries.

- **`broom`**: a part of the tidyverse and is designed to convert statistical analysis results into tidy data frames.

- **`forcats`**: designed to work with factors, which are used to represent categorical data. It simplifies the process of creating, modifying, and ordering factors.

- **`vcd`**: visualise and analyse categorical data.

**A useful shortcut to format your code: select all your code lines, use Ctrl+Shift+A for automatically format them in a tidy way.**

### Data overview

```{r,warning=FALSE}
if(!require("dplyr"))
  install.packages("dplyr",dependencies = T)
# Load necessary libraries 
if(!require("ggplot2"))
  install.packages("ggplot2",dependencies = T)
if(!require("broom"))
  install.packages("broom",dependencies = T)


library(dplyr) 
library(ggplot2)
library(broom)

```

Or we can use library `tidyverse` which includes `ggplot2`, `dplyr,broom` and other fundamental libraries together already, remember you need first install the package if you haven't by using `install.packages("tidyverse")`.

```{r,warning=FALSE}
if(!require("tidyverse"))
  install.packages("tidyverse",dependencies = T)

library(tidyverse)
```

We will also use 'forcats' library, so

```{r,warning=FALSE}
if(!require("forcats"))
  install.packages("forcats")

library(forcats)
```

Exactly as you did in previous weeks, we first load in the dataset:

```{r}
frs_data <- read.csv("../data/FRS/FRS16-17_labels.csv")
```

Recall in previous weeks, we used the following code to overview the dataset. Familiarise yourself again by using them:

```{r,results='hide'}
View(frs_data)
```

and also `summary()` to produce summaries of each variable

```{r,results='hide'}
summary(frs_data)
```

You may notice that for the numeric variables such as *hh_income_gross* (household gross income) and *work_hours*(worked hours per week), the `summary()` offers useful descriptive statistics. While for the qualitative information, such as *age_group* (age group), *highest_qual* (Highest educational qualification), *marital_status* (Marital status) and *nssec* (Socio-economic status), the `summary()` function is not that useful by providing mean or median values.

Performing descriptive analysis for categorical variables or qualitative variables, we focus on summarising the frequency and distribution of categories within the variable. This analysis helps understand the composition and diversity of categories in the data, which is especially useful for identifying patterns, common categories, or potential data imbalances.

```{r}
# Frequency count
table(frs_data$age_group)
table(frs_data$highest_qual)
table(frs_data$marital_status)
table(frs_data$nssec)
```

By using ggplot2, it is easy to create some nice descriptive charts for the categorical variables, just like what you did for the continuous variables last week.

```{r, warning=FALSE}
ggplot(frs_data, aes(x = highest_qual)) +
  geom_bar(fill="brown",width=0.5) +
  labs(title = "Histogram of Highest Qualification in FRS", x = "Highest Qualification", y = "Count")+#set text info
  theme_classic()#choose theme type, try theme_bw(), theme_minimal() see differences
```

```{r,warning=FALSE}

ggplot(frs_data, aes(x = health)) +
  geom_bar(fill="skyblue") +
  geom_text(stat = "count", aes(label = ..count..),vjust = -0.3,colour = "grey")+ #add text
  labs(title = "Histogram of Health in FRS", x = "Health", y = "Count")+#set text info
  theme_minimal()
```

```{r,warning=FALSE}
ggplot(frs_data, aes(x = nssec)) + 
  geom_bar(fill = "yellow4") + 
  labs(title = "Histogram of NSSEC in FRS", x = "NSSEC", y = "Count") +
  coord_flip()+ #Flip the Axes, add a # in front of this line, to make the code in gray and you will see why we would better flip the axes at here
  theme_bw() 
 
```

If we want to reorder the Y axis by from highest to lowest, we use the functions in `forcats` library. `fct_infreq()`: orders by the value's frequency of the variable `nssec`. `fct_rev()`: reverses the order to go from highest to lowest.

```{r}
ggplot(frs_data, aes(x = fct_rev(fct_infreq(nssec)))) + 
  geom_bar(fill = "yellow4") + 
  labs(title = "Histogram of NSSEC in FRS", x = "NSSEC", y = "Count") +
  coord_flip()+ #Flip the Axes, add a # in front of this line, to make the code in gray and you will see why we would better flip the axes at here
  theme_bw() 
 

```

You can change the variables in ggplot() to make your own histogram chart for the variables you are interested in. You will learn more of visualisation methods in Week 5's practical.

### Correlation

As before, rather than relying upon an impressionistic view of the strength of the association between two variables, we can measure that association by calculating the relevant correlation coefficients.

To calculate the correlation between categorical data, we first use Chi-squared test to assess the independence between pairs of categorical variables, then we use Cramer’s V to measures the strength of association - the correlation coefficents in R.

**Pearson’s chi-squared test** (χ2) is a statistical test applied to sets of categorical data to evaluate how likely it is that any observed difference between the sets arose by chance. If the p-value is low (typically \< 0.05), it suggests a significant association between the two variables.

```{r}
chisq.test(frs_data$health,frs_data$happy) 
```

If you see a warning message of Chi-squared approximation may be incorrect. This is because some expected frequencies in one or more cells of the cross-tabulation (health \* happy) are too low. The df means degrees of freedom and it relates to the size of the table and the number of categories in each variable. The most important message from the output is the estimated p-value, which shows as p-value \< 2.2e-16 (2.2 with 16 decimals move to the left, it is a very small number so written in scientific notation). P-value of the chi-squared test is far smaller than 0.05, so we can say the correlation is statistically significant.

**Cramér’s V** is a measure of association for categorical (nominal or ordinal) data. It ranges from 0 (no association) to 1 (strong association). The main downside of using Cramer’s V is that no information is provided on whether the correlation is positive or negative. This is not a problem if the variable pair includes a nominal variable but represents an information loss if both variables being correlated are ordinal.

```{r}
# Install the 'vcd' package if not installed 
if(!require("vcd"))   
install.packages("vcd", repos = "https://cran.r-project.org", dependencies = T)
library(vcd)  

# create the crosstable 
crosstab <- table(frs_data$health, frs_data$happy)

# Calculate Cramér's V 
assocstats(crosstab)

#you can also directly calculate the association between variables
assocstats(table(frs_data$health, frs_data$age_group))
```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
#### Q1: What are strongest and weakest associations health status? {.unnumbered}

**Which of our selected person-level variables is most strongly/weakly correlated with an individual’s health status?** Use the codes of Chi-square test and Cramer's V to answer this question by completing Table 1.
:::

**Table 1 Person-level correlations with health status**

|  |  |  |  |
|------------------|------------------|------------------|------------------|
| **Covariates** |  | **Correlation Coefficient** | **Statistical Significance** |
|  |  | *Cramer’s V* | *p-value* |
| *health* | *age_group* |  |  |
| *health* | *highest_qual* |  |  |
| *health* | *marital_status* |  |  |
| *health* | *nssec* |  |  |

## Income inequality with respect to gender and health status

In this section, we will work with individual-level data (“FRS16-17_label.csv”) to explore income inequality with respect to gender and health status.

To explore income inequality, we need to work with a data set excluding dependent children. In addition, we look at individuals who are the representative persons of households. Therefore, we will select cases (or samples) that meet both conditions.

We want R to select persons only if they are the representative persons of households and they are not dependent children. The involved variables are `hrp` and `Dependent` for the categories “Household Reference Person” and “independent”, you can select the appropriate cases. We also want to exclude the health variable reported as "Not known".

```{r}
frs_df <- frs_data %>% filter(hrp == "HRP" &
                                dependent == "Independent" &
                                health != "Not known") 
```

Then, we create a new numeric variable `Net_inc_perc` indicate net income per capita as our dependent variable:

```{r}
frs_df$Net_inc_pp = frs_df$hh_income_net / frs_df$hh_size  
summary(frs_df$Net_inc_pp)
```

The distribution of the net household income per capita can be visualised by using `ggplot()`

```{r}
ggplot(frs_df, aes(x = Net_inc_pp)) +
  geom_histogram(
    bins = 250, #A higher bins value means more, narrower bars, covers smaller range of value
    color = "black",
    fill = "skyblue",
    alpha = 0.7
  ) +   labs(title = "Distribution of Net_inc_pp", x = "Value", y =  "Frequency") +   scale_x_continuous(labels = scales::label_comma()) +  # Prevent scientific notation on the x axis
  theme_minimal()
```

The histogram shows that `Net_inc_pp` is **highly right-skewed**. Most households are concentrated in a narrow band between roughly £0 and £40,000 per person, with the peak around £10,000–£15,000, while a long tail stretches to the right up to a maximum of £864,812. This skewness is also visible in the `summary()` output: the mean (£15,834) is noticeably higher than the median (£13,347), because a small number of very high incomes pull the mean upwards. For skewed data like this, the median is a better measure of a "typical" value. There are also a few **negative values** (down to −£238,160), which can occur when households report losses, for example from self-employment. These extreme values are rare, but they stretch the x-axis so much that the main body of the distribution looks very thin. Keep them in mind when interpreting the regression results, as outliers can influence the estimated coefficients.

Now that we understand how net income per capita is distributed, we can ask **why** it varies so much between households. In this section, we explore whether income differs by **gender** and **health status**. If men and women, or people in better and worse health, have systematically different incomes, this would point to income inequality between these groups. Since both are categorical variables, we first need to check their categories before including them in a regression model.

Our two qualitative independent variables “`sex`” and “`health`”. Let's first know what they look like:

```{r}
table(frs_df$sex) 
table(frs_df$health)
```

Remember from the lecture what we did with Region in the long-term illness example before adding it to the regression model? Two steps are needed before including a categorical variable in a regression model:

1. **Convert it to a factor**, so that R treats it as categorical rather than as text or numbers.
2. **Choose the reference category** (baseline), as the model's coefficients will show how each of the other categories differs from it.

Here, we use **Female** and **Very Bad** health as the reference categories, but you can choose others if you wish. The code below combines both steps in one line: `as.factor()` converts the variable to a factor, and `fct_relevel()` sets the reference category.

```{r}
frs_df$sex <- fct_relevel(as.factor(frs_df$sex), "Female") 
frs_df$health <- fct_relevel(as.factor(frs_df$health), "Very Bad")
```

Implement the regression model with the two qualitative independent variables.

```{r}
model_frs <- lm(Net_inc_pp ~ sex + health, data = frs_df) 
summary(model_frs)
```

As in Week 2, here is what the code does:

**`lm()` Function**:

- `lm()` stands for "linear model" and is used to fit a linear regression model in R.
- The formula syntax `Net_inc_pp ~ sex + health` specifies a relationship between:
  - **Dependent Variable**: `Net_inc_pp`.
  - **Independent Variables**: `sex` and `health`. The model is fitted using the `frs_df` dataset (`data = frs_df`).

**Storing the Model**: The `model_frs <-` syntax stores the fitted model in an object called `model_frs`.

`summary(model_frs)` provides a detailed output of the model's results, including:

- **Coefficients**: Estimates of how each category differs from its reference category in `Net_inc_pp` (i.e., how each independent variable is associated with `Net_inc_pp`).
- **Standard Errors**: How precise each coefficient estimate is (smaller = more precise).
- **t-values** and **p-values**: Indicate the statistical significance of the effect of each independent (explanatory) variable.
- **R-squared** and **Adjusted R-squared**: Show how well the independent variables explain the variance in the dependent variable.
- **F-statistic**: Tests the overall significance of the model.

The result can be formatted by:

```{r warning=FALSE}
library(broom) 
tidy(model_frs)
```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
### Q2. Income inequality by gender and health {.unnumbered}

**What conclusions could be drawn in terms of income inequalities with respect to gender and health status? Also think about the statistical significance of these differences.**
:::

The interpretation of this table is again similar to what you have learnt in Week 2, but you may find that R has automatically treated the categorical/qualitative variables **sex** and **health** as dummy variables. Each category gets a coefficient except the reference category set by `fct_relevel()`. That is why the output has `sexMale` but no `sexFemale`, and no `healthVery Bad`.

Each coefficient is therefore interpreted as a difference from the reference category, **Female** with **Very Bad** health, holding the other variable constant:

- The **Intercept** (**£12,116**) is the predicted net income per capita for the reference group: females with Very Bad health.

- The association of **`sexMale`** is positive and statistically significant (**+£2,091, p \< 0.001**): men have £2,091 more net income per capita than women with the same health status, indicating a gender gap in income.

- The associations of **`healthBad`** and **`healthFair`** are not statistically significant (**p = 0.90** and **0.18**): there is no evidence that their net income per capita differs from people in Very Bad health.

- The associations of **`healthGood`** and **`healthVery Good`** are positive and statistically significant (**+£2,766** and **+£4,932**, both **p \< 0.001**): compared with people in Very Bad health, people in Good and Very Good health have £2,766 and £4,932 more net income per capita respectively, suggesting that income rises with better health.

We then read R-squared and Adjusted R-squared to evaluate the model fit. Both are only around **0.016**, meaning `sex` and `health` together explain only 1.6% of the variation in `Net_inc_pp`. The model is therefore poor for prediction, and more variables (e.g. education level, occupation, age) would be needed to explain income better. Even so, the F-test is highly significant (**p \< 2.2e-16**), so the income gaps by gender and health identified above are real, even though they explain only a small part of why incomes differ.

## **Implementing a linear regression model with a qualitative independent variable**

To gain further practice on using linear regression models with qualitative variables as independent variables, we extend the model you created last week by adding the qualitative/categorical variable Region from the Census 2021 dataset. This allows us to answer **Research Question 2: How does health vary across regions in England and Wales?** You should already have the dataset **EW_DistrictPercentages.csv**; if not, you can download it from [Canvas](https://canvas.liverpool.ac.uk/courses/93312/pages/census-data?module_item_id=2727713). Please put the .csv file in your '/data/Census2021' folder.

The practical is split into two main parts. The first focuses on implementing a linear regression model with a qualitative independent variable. **Note that** you need first to set the reference category (baseline) as the outcomes of the model reflects the **differences** between categories with the baseline. The second part focuses on prediction based the estimated linear regression model.

First we load the UK district-level census dataset.

```{r warning=FALSE}
# load data
LAcensus <- read.csv("../data/Census2021/EW_DistrictPercentages.csv") # Local authority level
```

Using the district-level census dataset "**EW\_\_DistrictPercentages.csv**". the variable "Region" (labelled as Government Office Region) is used to explore regional inequality in health.

Familiarise yourself with the dataset by using the same codes as last week:

```{r,results='hide'}
#view the data 
View(LAcensus)  
glimpse(LAcensus)
```

The names() function returns all the column names.

```{r,results='hide'}
names(LAcensus)
```

The dim() function simply returns the number of rows and number of columns.

```{r}
dim(LAcensus) 
```

There are 331 rows and 151 columns in the dataset. It would be very hard to scan through the data if we use so many variables altogether. Therefore, we can select several columns to tailor for this practical. You can of course include other variables you are interested in also by their names:

```{r}
df <- LAcensus %>% select(c("pct_Longterm_sick_or_disabled",
                            "pct_No_qualifications",
                            "pct_Males",
                            "pct_Higher_manager_prof",
                            "Region"))
```

Simply descriptive of this new data

```{r}
summary(df)
```

Now we can retrieve the "Region" column from the data frame by simply use **df\$Region**. But what if we want to understand the data better, like the following questions?

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
### Q3. Categories of Region {.unnumbered}

**How many categories does the variable "Region" entail? How many local authority districts does each region include?**
:::

Simply use the function table() would return you the answer.

```{r}
table(df$Region) 
```

The `table()` function tells us that this data frame contains 10 regions, and the number of LAs belongs to each region.

R can only include the categorical variables in the **factor** type, so we set the column *Region* in `factor()`

```{r}
df$Region<- factor(df$Region) 
```

### **Include the categorical variables into a regression model**

We will continue with a very similar regression model fitted in last week that relates Percentages long-term illness (*pct_Long_term_sick_or_disabled*) to Percentages no-qualification (*pct_No_qualifications*), Percentage Males (*pct_Males*) and Percentages Higher Managerial or Professional occupation (*pct_Higher_manager_prof*).

Decide which region to be set as the baseline category. The principle is that if you want to compare the (average) long term illness outcome of Region A to those of other regions, Region A should be chosen as the baseline category. For example, if you want to compare the (average) long term illness outcome of London to rest of regions in the England and Wales, London should be selected as the baseline category.

Implement the regression model with the categorical variables - *Region* in our case. R will automatically handle the qualitative variable as dummy variables so you don't need to concern any of that. But you need to let R knows which category of your qualitative variable is your reference category or the baseline. Here we will use London as our first go. **Note:** We choose London as the baseline category so the London region will be [**excluded**]{.underline} in the independent variable list.

Therefore, first, we set London as the reference:

```{r}
df$Region <- fct_relevel(df$Region,  "London")
```

Similar to last week, we build our linear regression model, but also include the *Region* variable into the model.

```{r}
model <- lm(pct_Longterm_sick_or_disabled ~ pct_Males + pct_No_qualifications + pct_Higher_manager_prof + Region, data = df)

summary(model)
```

You have already learnt how to interpret the output of regression model last week: **Significance** (p-value), **Coefficient Estimates, and Model fit** (R squared and Adjusted R-squared).

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
#### Q4. Regional differences compared with London {.unnumbered}

**Relating back to this week's lecture notes, indicate what regions have statistically significant differences in the percentage of long-term illness, compared to London?**
:::

First, the **Significance** and the **Coefficient Estimates**. The p-values in the last column of the output table show that most of the independent variables are significant predictors of `pct_Longterm_sick_or_disabled`. All coefficients below are interpreted **when the other predictors remain the same**.

- The **Intercept** (**7.62**) is the predicted long-term sick/disabled rate for a **London** district (the reference category) when `pct_Males`, `pct_No_qualifications` and `pct_Higher_manager_prof` are all 0%. Since no real district has 0% males, we do not interpret this value directly; it is the baseline value to which the other coefficients are added when making predictions.

- **Continuous variables.** Similar to last week, `pct_No_qualifications` is significant at the \<0.001 level (three asterisks \*\*\*), indicating a highly statistically significant relationship. We are slightly less confident about `pct_Males`, which is significant at the \<0.01 level (two asterisks \*\*). Their coefficient estimates tell us that:

  - The association of **`pct_No_qualifications`** is positive and statistically significant (**+0.25, p \< 0.001**): each 1% increase in residents with no qualifications is associated with a 0.25% increase in the long-term sick/disabled rate, when the other predictors remain the same.

  - The association of **`pct_Males`** is negative and statistically significant (**−0.17, p \< 0.01**): each 1% increase in males is associated with a 0.17% decrease in the long-term sick/disabled rate, when the other predictors remain the same.

  - The association of **`pct_Higher_manager_prof`** is not statistically significant (**p = 0.72**).

- **Region dummy variables.** R has created a dummy variable (the items starting with `Region`) for every region except the reference category, **London**. Each coefficient therefore shows how much higher (positive) or lower (negative) a region's long-term sick/disabled rate is **compared with London**, when the other predictors remain the same:

  - **Significant at the \<0.001 level (\*\*\*):** `RegionEast`, `RegionNorth East`, `RegionNorth West` and `RegionWales`. For example, the North East has a **1.16%** higher rate than London, and Wales a **1.38%** higher rate, while the East has a **0.76%** lower rate.

  - **Significant at the 0.05 level (\*):** `RegionSouth East` and `RegionWest Midlands`. This is weaker evidence than \*\*\*, so they are only mildly statistically significant.

  - **Not significant:** `RegionEast Midlands`, `RegionSouth West` and `RegionYorkshire and The Humber`, meaning their rates are not significantly different from London's.

  You can interpret the other regions in the same way using their coefficient estimates.

  **Reminder**: You **cannot** tell from this output whether North East and Wales, or any two regions other than London, are significantly different from each other, because each region's coefficient only compares that region with London. To compare North East and Wales, set either of them as the reference category, using `df$Region <- fct_relevel(df$Region, "North East")` or `df$Region <- fct_relevel(df$Region, "Wales")`, and re-run the model.

Last but not least, the **Model Fit**. R-squared (**0.759**) and Adjusted R-squared (**0.750**) are both around 0.75: the model explains 75.9% of the variation in the long-term sick/disabled rate across districts, or 75.0% after adjusting for the number of independent variables. Together, they suggest a strong fit of the model.

### **Change the baseline category**

If you would like to learn about differences in long-term illness between North East and other regions in the EW, you need to change the baseline category (from London) to the North East region (with variable name “Region_2”).

```{r}
df$Region <- fct_relevel(df$Region, "North West")
```

The regression model is specified again as follows:

```{r}
model1 <- lm(
  pct_Longterm_sick_or_disabled ~ pct_Males + pct_No_qualifications + pct_Higher_manager_prof + Region,
  data = df
)

summary(model1)
```

Examining R-squared to measure the model fit. How robust is the new model? What % of the variance in the dependent variable has been explained?

Now, complete the following table.

| Region names | Higher or lower than North West | Whether the difference is statistically significant (Yes or No) |
|:----------------------:|:----------------------:|:----------------------:|
| East Midlands |  |  |
| East |  |  |
| North East |  |  |
| South East |  |  |
| London |  |  |
| West Midlands |  |  |
| Yorkshire and The Humber |  |  |
| Wales |  |  |

Check the answer at the end of this practical page.

Now it is very easy to use your model to estimate the results Y (dependent variable - `pct_Longterm_sick_or_disabled` ) by setting all the input independent variable X (`pct_Males` `pct_No_qualifications` and `pct_Higher_manager_prof`).

```{r}
obj_London <- data.frame(
  pct_Males = 49.7,
  pct_No_qualifications = 24.3,
  pct_Higher_manager_prof = 14.7,
  Region = "London"
) 
obj_WM <- data.frame(
  pct_Males = 49.8,
  pct_No_qualifications = 23.3,
  pct_Higher_manager_prof = 11.2,
  Region = "West Midlands"
) 
obj_NE <- data.frame(
  pct_Males = 49.8,
  pct_No_qualifications = 23.3,
  pct_Higher_manager_prof = 11.2,
  Region = "North East"
  
) 

predict(model1, obj_London) 
predict(model1, obj_WM) 
predict(model1, obj_NE)
```

### Recode the Region variable and explore regional inequality in health

In many real-world studies, we might not be interested in health inequality across all regions. For example, in this case study, we are interested in health inequality between *London*, *South, North, Midland and Wales*. We can achieve this by re-grouping regions in England and Wales based on the variable “Region”. That said, we need to have a new grouping of regions as follows:

|                            |                       |
|----------------------------|-----------------------|
| **Original region labels** | **New region labels** |
| East Midlands              | Midlands              |
| East                       | South                 |
| London                     | London                |
| North East                 | North                 |
| North West                 | North                 |
| South East                 | South                 |
| South West                 | South                 |
| West Midlands              | Midlands              |
| Yorkshire and The Humber   | North                 |
| Wales                      | Wales                 |

Here we use mutate() function in R to make it happen:

```{r}
df <- df %>%
  mutate(
    New_region_label = fct_collapse(
      Region,
      North = c("North East", "North West", "Yorkshire and The Humber"),
      Midlands = c("East Midlands", "West Midlands"),
      South = c("East", "South East", "South West"),
      London = "London",
      Wales = "Wales"
    )
  )
```

This code may look a bit complex. You can simply type ?mutate in your console. Now in your right hand Help window, the R studio offers you the explanation of the mutate function. This is a common way you can use R studio to help you learn what the function ca`ate()` creates new columns that are functions of existing variables. Therefore, the `df %>% mutate()` means add a new column into the current dataframe `df`; the `New_region_label` in the `mutate()` function indicates the name of this new column is `New_region_label`. The right side of the `New_region_label =` indicates the value we want to assign to the `New_region_label` in each row.

The right side of `New_region_label` is

`fct_collapse(Region, North = c("North East", "North West", "Yorkshire and The Humber"), Midlands = c("East Midlands", "West Midlands"), South = c("East", "South East", "South West"), London = "London", Wales = "Wales")`

By using the code, the `fct_collapse()` function recodes or group each value in the `Region` column into one of the broader categories: "London", "South", "North","Midlands" or "Wales". Specifically, it takes the existing factor levels in `Region` and collapses multiple detailed regions into fewer, aggregated groups. For example, regions such as *"North East"*, *"North West"*, and *"Yorkshire and The Humber"* are all grouped under **"North"**, while *"East Midlands"* and *"West Midlands"* are combined into **"Midlands"**, and so on.

Now we use the same way to examine our new column `New_region_label`:

```{r}
table(df$New_region_label)
```

Comparing with the `Region_label`, we now can see the mutate worked:

```{r,results='hide'}
df[,c("Region","New_region_label")]
```

Now you will have a new qualitative variable named `New_region_label` in which the UK is divided into five regions: London, South, North, Midlands and Wales.

*Based on the newly generated qualitative variable* `New_region_label`, we can now build our new linear regression model. Don't forget:

\(1\) R need to deal with the categorical variables in regression model in the factor type;

```{r}
class(df$New_region_label)
```

The `class()` returns the type of the variable. The `New_region_label` is already a factor variable. If not, we need to convert it by the `as.factor()`, as we used above.

```{r}
df$New_region_label = as.factor(df$New_region_label)
```

\(1\) Let R know which region you want to use as the baseline category. Here I will use London again, but of course you can choose other regions.

```{r}
df$New_region_label <- fct_relevel(df$New_region_label, "London")
```

The linear regression window is set up below. This time we include `New_region_label` rather than `Region_label` as the region variable:

```{r}
model2 <- lm(
  pct_Longterm_sick_or_disabled ~ pct_Males + pct_No_qualifications + pct_Higher_manager_prof + New_region_label,
  data = df
)

summary(model2)
```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
#### Q5. London vs North, and South vs Wales {.unnumbered}

**Are there statistically significant differences in the percentage of people with long-term sick/disable between London and North, and between South and Wales, controlling for other variables? What conclusions could be drawn in terms of regional differences in health outcome?**
:::

## Predictions using fitted regression model

**Write down the % long term sick or disabled regression model with the new region label categorical variables.**

Relating to this week’s lecture, the % `pct_Longterm_sick_or_disabled` is equal to:

**\[write down the model\]**

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
### Q6. Predicting long-term illness in Wales and London {.unnumbered}

**Now imagine that the values of variables *pct_Males, pct_No_qualifications, and pct_Higher_manager_prof* are 49, 23 and 11, respectively, what would the percentage of persons with long-term sick/disable in Wales and London be?**
:::

Check the answer at the end of this practical page.

## **Answers to the Questions**

### Q1 associations with health status

Using the Chi-squared test and Cramér's V, Table 1 can be completed as follows:

| **Covariates** |                  | **Cramér's V** | **p-value** |
|:--------------:|:----------------:|:--------------:|:-----------:|
|    *health*    |   *age_group*    |     0.329      |  \< 0.001   |
|    *health*    |  *highest_qual*  |     0.287      |  \< 0.001   |
|    *health*    |     *nssec*      |     0.280      |  \< 0.001   |
|    *health*    | *marital_status* |     0.229      |  \< 0.001   |

All four associations are statistically significant (p \< 0.001). **Age group** has the strongest association with health status (Cramér's V = 0.329), and **marital status** the weakest (0.229). Remember that Cramér's V tells us the strength of an association, but not its direction.

### Q2: income inequality by gender and health status

- **Gender:** men have £2,091 more net income per capita than women with the same health status (p \< 0.001), indicating a significant gender gap in income.
- **Health:** compared with people in Very Bad health, people in Good and Very Good health have £2,766 and £4,932 more net income per capita respectively (both p \< 0.001). The differences for Bad and Fair health are not statistically significant.
- **Model fit:** sex and health explain only 1.6% of the variation in income (R² = 0.016), so other factors matter much more. However, the income gaps by gender and health are statistically significant.

### Q3: categories of Region

`table(df$Region)` shows **10 regions** with 331 local authority districts in total: East (45), East Midlands (35), London (33), North East (12), North West (39), South East (64), South West (30), Wales (22), West Midlands (30) and Yorkshire and The Humber (21).

### Q4: regions that differ significantly from London

Holding the other predictors constant, compared with London:

- **Significantly higher:** Wales (+1.38%), North East (+1.16%) and North West (+0.78%), all p \< 0.001.
- **Significantly lower:** East (−0.76%, p \< 0.001), West Midlands (−0.42%, p \< 0.05) and South East (−0.33%, p \< 0.05).
- **Not significantly different:** East Midlands, South West, and Yorkshire and The Humber.

### Changing the baseline to North West

R-squared (0.759) and Adjusted R-squared (0.750) are **exactly the same** as in the model with London as the baseline. Changing the reference category does not change the model fit; it only changes which comparisons the Region coefficients show.

| Region names | Higher or lower than North West | Statistically significant? |
|:----------------------:|:----------------------:|:----------------------:|
| East Midlands | Lower (−1.04%) | Yes |
| East | Lower (−1.55%) | Yes |
| North East | Higher (+0.38%) | No (p = 0.11) |
| South East | Lower (−1.11%) | Yes |
| South West | Lower (−0.69%) | Yes |
| London | Lower (−0.78%) | Yes |
| West Midlands | Lower (−1.20%) | Yes |
| Yorkshire and The Humber | Lower (−0.90%) | Yes |
| Wales | Higher (+0.60%) | Yes |

Note that London is 0.78% lower than North West, which is the same size as the North West coefficient in the London-baseline model (+0.78%), with the opposite sign.

### Q5: London vs North, and South vs Wales

- **London vs North:** yes. With London as the baseline, the North has a 0.47% higher long-term sick/disabled rate than London (p = 0.012), when the other predictors remain the same.

- **South vs Wales:** this **cannot** be answered from `model2`, because both South and Wales are compared only with London. To test it, set one of them as the reference category, e.g. `df$New_region_label <- fct_relevel(df$New_region_label, "South")`, and re-run the model. Doing so shows that Wales has a 1.76% higher rate than the South (p \< 0.001).

- **Conclusion:** after controlling for sex, qualifications and occupation, `model2` tells us how each region compares with **London**: Wales (+1.25%) and the North (+0.47%) have significantly higher long-term sick/disabled rates than London, while the South (−0.51%) and the Midlands (−0.41%) have significantly lower rates. This suggests clear regional inequalities in health across England and Wales.

  **Reminder**: these conclusions are all *relative to London*. Although the coefficients suggest an ordering (e.g. Wales appears higher than the North, and the North higher than the South), `model2` **cannot** tell us whether the differences between two non-London regions are statistically significant. To test any such comparison, the model must be **re-built** with one of those regions as the reference category, as we did above for South vs Wales.

### The written-down model and Q6

The model of the new region label is: pct_Longterm_sick_or_disabled (%) = 9.98+ (-0.1842)\* pct_Males (%) + 0.2014 \* pct_No_qualifications (%) + (-0.0341) \* pct_Higher_manager_prof + (0.4742) \* North + (-0.5063) \* South + (-0.4103) \* Midlands + 1.2518 \* Wales, where North, South, Midlands and Wales are dummy variables (1 if the district is in that region, 0 otherwise). London is the reference category, so all four dummies are 0 for London. Note that `pct_Higher_manager_prof` is not statistically significant at the 0.05 level (p = 0.087).

If `pct_Males`, `pct_No_qualifications` and `pct_Higher_manager_prof` are 49, 23 and 11:

- **Wales** (Wales = 1, other dummies = 0): `9.98 + (-0.1842)*49 + 0.2014*23 + (-0.0341)*11 + 1.2518*1`
- **London** (all dummies = 0): `9.98 + (-0.1842)*49 + 0.2014*23 + (-0.0341)*11`

You can paste these directly into your RStudio Console to get the results. You can also make a new object and use `predict()`:

```{r}
obj_Wales <- data.frame(
  pct_Males = 49,
  pct_No_qualifications = 23,
  pct_Higher_manager_prof = 11,
  New_region_label = "Wales"
)

obj_London <- data.frame(
  pct_Males = 49,
  pct_No_qualifications = 23,
  pct_Higher_manager_prof = 11,
  New_region_label = "London"
)

predict(model2, obj_Wales)
predict(model2, obj_London)
```

**Therefore, the predicted percentages of persons with long-term illness in Wales and London are 6.47% and 5.22% respectively.** The difference (1.25%) is exactly the Wales coefficient, because the two districts have the same values for all the other predictors. If you got the right answers, congratulations: you can now use a regression model to make predictions!