---
title: "Lab: Logistic Regression"
author: "Zi Ye"
date: "`r Sys.Date()`"
output: html_document
---

Last week, we learnt how to use qualitative variables in a multiple linear regression model to understand the relationship between independent variables X1 ... Xn and the dependent variable Y. Today we will learn to use logistic regression for a binary dependent variable. A **logistic regression model** is a type of regression analysis used when the dependent variable is binary (e.g., "success/failure" or "0/1"). It estimates the probability of one outcome relative to the other using a logistic function. This model is commonly used in situations like predicting disease presence (yes/no) or customer churn (stay/leave). The independent variables can be continuous, categorical, or a mix of both. The model's output is in the form of odds ratios, showing how predictors affect the likelihood of the outcome.

The lecture’s slides can be found [here](https://github.com/GDSL-UL/stats/blob/main/lectures/lecture04.pdf).

**Learning objectives**

An understanding of, and ability to:

-   Estimate and interpret a logistic regression model

-   Assess the model fit

The application context of a binomial logistic regression model is when the dependent variable under investigation is a binary variable. Usually, a value of 1 for this dependent variable means the occurrence of an event; and, 0 otherwise. For example, the dependent variable for this practical is whether a person is a long-distance commuter i.e. 1, and 0 otherwise.

In this week’s practical, we are going to apply logistic regression analysis in an attempt to answer the following research question:

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**RESEARCH QUESTION: Who is willing to commute long distances?**
:::

The practical is split into two main parts. The first focuses on implementing a binary logistic regression model with R. The second part focuses on the interpretation of the resulting estimates.

## Knowing the dataset and descriptive analysis

Prepare the data for implementing a logistic regression model. The data set used in this practical is the “sar_sample_label.csv” and "sar_sample_code.csv". The SAR is a snapshot of census microdata, which are individual level data. The data sample has been drawn and anonymised from census and known as the Samples of Anonymised Records (SARs).

You may need to download the two datasets and also the data dictionary from [Canvas](https://canvas.liverpool.ac.uk/courses/93312/pages/sample-of-anonymised-records-sar-dataset) if you haven't. Double click to open both .csv files you may find they have exactly same column names - yes, they are in fact the same spreadsheet but 'sar_sample_label.csv' is more readable for the cell contents as they are in text format but cells in the 'sar_sample_code.csv' is in coding format. The 'SAR_dictionary' is created to help you understand what the columns mean, just like the FRS dictionary for the FRS datasets last week. Yes, the two SAR csv files are actually the same dataframe, only one uses the label as the value but the other uses the code. This is quite normal: when we do surveys, we use text as the options of multiple choice questions for respondents to choose from, and therefore the collected survey results are in the 'sar_sample_label' format. The label format is usually more readable and is good to do descriptive analysis as we will show in today's content. However, we may find the coding format is more concise for writing the code and thus will be used during regression analysis. **Please notice that coding the labels into numbers, doesn't mean the categorical/qualitative variable has been converted into numeric/continuous numbers.**

Let's first read in both for the data overview.

```{r, message=FALSE}
if(!require("tidyverse"))
  install.packages("tidyverse",dependencies = T, repos = "https://cloud.r-project.org/")
if(!require("broom"))
  install.packages("broom",dependencies = T, repos = "https://cloud.r-project.org/")
if(!require("forcats"))
  install.packages("forcats")

library(tidyverse)
library(broom)
library(forcats)
```

```{r,results='hide'}
sar_label <- read.csv("../data/SAR/sar_sample_label.csv")
sar_code <- read.csv("../data/SAR/sar_sample_code.csv")
```

Run the following codes to view both dataframes:

```{r,results='hide'}
#View the sar_label dataset
View(sar_label)
```

```{r,results='hide'}
#View the sar_code dataset
View(sar_code)
```

To know the dataset better, we can run dim() to know how many individuals (rows) in this sample data and how many variables (columns) have been recorded.

```{r}
dim(sar_label)
#or
dim(sar_code)
```

So there are 50,000 respondents in the sample dataset and 39 variables are included.

After browsing both data sheets and also the data dictionary (sometimes we call it meta data), you may notice that all the columns in the SAR dataset are of categorical/qualitative type. Therefore, if we run the `summary()` function for the dataframes, as we did in previous weeks for the Census data or the FRS dataset, the outputs of the `summary()` may not be very useful. Again, please notice that the figures in sar_code.csv are not actual numbers, but codes for the different categories of the qualitative variables. The `summary()` results here are therefore not useful. Think back to what we learnt last week: when the variables are of categorical/qualitative type, how do we carry out the descriptive analysis for them?

```{r,results='hide'}
summary(sar_code)
summary(sar_label)
```

So, the answer is we focus on summarising the frequency and distribution of categories within the variable and the function in R for doing this is `table()`. For example, the variable “work_distance” captures a person’s commuting distance. We can run the table() for both label and code csv to understand the distribution, composition and diversity of categories of “work_distance” in the data. Try to repeat what we have learnt in Week 3 to finish the **descriptive analysis** for the categorical variables in SAR (by ggplot2).

```{r}
table(sar_label$work_distance)
table(sar_code$work_distance)
```

To create a chart as part of the descriptive analysis, exactly as what we did in Week 3, we use the library `ggplot2` and the code:

```{r}
library(ggplot2)
ggplot(sar_label, aes(x = work_distance)) + 
  geom_bar(fill = "rosybrown") + 
  labs(title = "Histogram of work distance in SAR sample", x = "Work Distance", y = "Count") +
  coord_flip()+ #Flip the Axes, add a # in front of this line, to make the code in gray and you will see why we would better flip the axes at here
  theme_bw() 
```

We have learnt that for the categorical/qualitative variables, we need R to treat them as `factor` type. Therefore, if we want to create descriptive analysis charts by using `sar_code` , we first need to convert the code numbers from numeric to factor type.

```{r}
library(ggplot2)
ggplot(sar_code, aes(x = as.factor(work_distance))) + 
  geom_bar(fill = "rosybrown") + 
  labs(title = "Histogram of work distance in SAR sample", x = "Work Distance", y = "Count") +
  coord_flip()+ #Flip the Axes, add a # in front of this line, to make the code in gray and you will see why we would better flip the axes at here
  theme_bw() 
```

You may also realise that this chart is not that readable to be used as a descriptive analysis chart to put in the report as the Y-axis labels are all codes, therefore the first chart created by labels is a better choice. Based on 'SAR_dictionary.xlsx', we learn how the codes map to the categories for the variable work_distance in both csv.

| Code for Work_distance | Categories                                   |
|------------------------|----------------------------------------------|
| -9                     | Age\<16 or not working                       |
| 1                      | Less than 2 km                               |
| 2                      | 2 to \<5 km                                  |
| 3                      | 5 to \<10 km                                 |
| 4                      | 10 to \<20 km                                |
| 5                      | 20 to \<40 km                                |
| 6                      | 40 to \<60 km                                |
| 7                      | 60km or more                                 |
| 8                      | At home                                      |
| 9                      | No fixed place                               |
| 10                     | Work outside England and Wales but within UK |
| 11                     | Work outside UK                              |
| 12                     | Works at offshore installation (within UK)   |

There are a variety of categories in the variable, however, we are only interested in commuting distance and therefore in people reporting their commuting distance. **Thus, we will explore the numeric codes of the variable ranging from 1 to 8.**

As we are also interested in exploring whether people with different socio-economic statuses (or occupations) tend to be associated with varying probabilities of commuting over long distances, we further filter or select cases.

```{r}
table(sar_label$nssec)
table(sar_code$nssec)
```

```{r}
ggplot(sar_label, aes(x = fct_rev(fct_infreq(nssec)))) + 
  geom_bar(fill = "tan") + 
  labs(title = "Histogram of NSSEC in SAR sample", x = "NSSEC", y = "Count") +
  coord_flip()+ #Flip the Axes, add a # in front of this line, to make the code in gray and you will see why we would better flip the axes at here
  theme_bw() 
```

From 'SAR_dictionary.xlsx', we learn what each code indicate corresponding category for variable `nssec`. For the following regression model, we select people who reported an occupation, and delete cases with numeric codes from 9 to 12, who are *unemployed*, *full-time students*, *children* and *not classifiable*.

| Code for nssec | Category labels                               |
|----------------|-----------------------------------------------|
| 1              | Large employers and higher managers           |
| 2              | Higher professional occupations               |
| 3              | Lower managerial and professional occupations |
| 4              | Intermediate occupations                      |
| 5              | Small employers and own account workers       |
| 6              | Lower supervisory and technical occupations   |
| 7              | Semi-routine occupations                      |
| 8              | Routine occupations                           |
| 9              | Never worked or long-term unemployed          |
| 10             | Full-time student                             |
| 11             | Not classifiable                              |
| 12             | Child aged 0-15                               |

Now, as we did last week, we use `filter()` to prepare our dataframe. You may already realise that using `sar_code` is easier to do the filtering.

```{r}
sar_df <- sar_code %>% filter(work_distance<=8 & nssec <=8 )
```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**Q1. Create descriptive analysis for the two variables “work_distance”, “nssec” and "sex" with the new data, including (1) summarise the frequencies and (2) create histogram charts for the three variables.**
:::

## Preparing the input variables

For a logistic regression model, we first need to **recode the “work_distance” variable into a binary dependent variable**.

A simple way to create a binary dependent variable representing long-distance commuting is to use the mutate() function as discussed in last week's practical session. Before creating the binary variables from the “work_distance” variable, we need to define *what counts as a long-distance commuting move*. Such definition can vary. Here we define a long-distance commuting move as any commuting move over a distance above 60km (the category of “60km or more”).

```{r}
sar_df <- sar_df %>% mutate(
  New_work_distance = if_else(work_distance >6, 1,0))

```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**Q2.** **Do descriptive analysis on the new `sar_df` dataframe with new column named `New_work_distance` by using the codes you have learnt.**
:::

The independent variables are gender and socio-economic status. The gender variable `sex` is a categorical variable. Therefore, as we've learnt last week, before adding the categorical variables into the regression model, we first need to make it a factor and then identify the reference category. For gender, we use male as the baseline. Prepare your "sex" variable before the regression model:

```{r}
sar_df$sex <- relevel(as.factor(sar_df$sex),ref="1")
```

Then, we prepare socio-economic status variable `nssec` for the regression model. We are interested in whether people with occupations being "Higher professional occupations” are associated with a lower probability of commuting over long distances when comparing to people in other occupations. Therefore by using the "Higher professional occupations" code 2, we run the code as below:

```{r}
sar_df$nssec <- relevel(as.factor(sar_df$nssec), ref = "2")
```

## **Implementing a logistic regression model**

The binary dependent variable is long-distance commuting, variable name `New_work_distance`.

```{r}
#create the model
m.glm = glm(New_work_distance~sex + nssec, 
            data = sar_df, 
            family= "binomial")
# inspect the results
summary(m.glm) 
```

```{r}
# odds ratios
exp(coef(m.glm)) 
```

```{r}
# confidence intervals
exp(confint(m.glm, level = 0.95)) 
```

### **Interpreting estimated regression coefficients**

Compared to statistical inference in a multiple linear regression model (which you did in Weeks 2 and 3), the interpretation of coefficients (B) and odds ratios (Exp(B)) for the logistic regression model has some similarities and some differences:

**It is the same that we read p-values of regression coefficients to assess significances of coefficients**; for instance, by comparing p-values to the conventional level of significance of 0.05:

·       If the p-value of a coefficient is smaller than 0.05, the coefficient is statistically significant. In this case, you can say that the relationship between an independent variable and the outcome variable is *statistically* significant.

·       If the p-value of a coefficient is larger than 0.05, the coefficient is statistically insignificant. In this case, you can say or conclude that there is no statistically significant association or relationship between an independent variable and the outcome variable.

**It is different in the way we interpret the regression coefficients (B) as we need to read the odds ratios (Exp(B)) from `exp(coef(m.glm))`**

o   For the variable **`sex`**, the negative coefficient and the odds ratio indicate that the odds of commuting over long distances for a female are 0.693 times those of a male (the reference group), with a confidence interval (CI), or likely range, of 0.63 to 0.75, holding all other variables constant (the socio-economic classification variable). Put differently, being female reduces the odds of long-distance commuting by 30.7% (1-0.693).

o   For the variable **`nssec`**, the coefficients and odds ratios indicate the following for those whose socio-economic classification is:

-   the p-value of Large employers and higher managers (nssec=1) is \> 0.05, so there is no statistically significant relationship between large employers and higher managers and long-distance commuting;

-   Lower managerial and professional occupations (nssec=3) have 0.679 times the odds of long-distance commuting compared to our reference category (nssec=2, higher professional occupations), holding all other variables constant (the sex variable), with a likely range (CI) of 0.60 to 0.77. Therefore we can also say that workers in lower managerial and professional occupations have 32.1% (1-0.679) lower odds than higher professional workers of travelling more than 60km for work, for the same gender.

-   Small employers and own account workers (nssec=5) have 3.409 times the odds of the reference category (nssec=2, higher professional occupations), holding the gender variable constant, with a likely range (CI) of 3.00 to 3.87.

-   Compared to higher professional occupations (nssec=2), workers in routine occupations (nssec=8) have 0.226 times the odds of travelling more than 60km to work, with a CI of 0.19 to 0.27, when the other variables are held constant. Equivalently, being in a routine occupation reduces the odds of long-distance commuting by 77.4% (1-0.226).

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**Q3**. **Can you write down the findings you learnt from the model outcomes to other occupation groups (nssec = 4, nssec = 6 and nssec =7)?**
:::

### **Model fit**

We include the R library `pscl` to calculate the measures of fit.

```{r}
if(!require("pscl"))
  install.packages("pscl", repos = "https://cloud.r-project.org/")
library(pscl)
```

Relating back to this week’s lecture notes, what is the Pseudo R^2^ of the fitted logistic model (from the Model Summary table below)?

```{r,warning=FALSE}
# Pseudo R-squared
pR2(m.glm)

# or in better format
pR2(m.glm) %>% round(4) %>% tidy()
```

-   **llh**: The log-likelihood of the fitted model.

-   **llhNull**: The log-likelihood of the null model (without predictors).

-   **G2**: The likelihood ratio statistic, showing the model’s improvement over the null model.

-   **McFadden**: McFadden’s pseudo R-squared (a common measure of model fit).

-   **r2ML**: Maximum likelihood pseudo R-squared.

-   **r2CU**: Cox & Snell pseudo R-squared.

In multiple linear regression, R-squared indicates the % of the variance in the dependent variable that is explained by the independent variables. In a logistic regression model, R-squared is not directly applicable. Instead, we use pseudo R-squared measures, such as McFadden’s pseudo R-squared, or Cox & Snell pseudo R-squared to provide an indication of model fit. For an individual-level dataset like the SAR, a value around 0.3 is considered a good fit. Our model is therefore not that robust for prediction, but it still describes the association between the variables and the categories. To improve the model, we may need more variables as the independent variables, you may identify some based on the related literature or debates on this topic.

### Recode Socio-economic status variable and explore commuting differences

This time, we want to know whether "Lower supervisory and technical occupations", "Semi-routine occupations" and "Routine occupations" are associated with a higher probability of commuting over long distances compared to people in other occupations.

We can use mutate() to create a new column, set the value of "Lower supervisory and technical occupations", "Semi-routine occupations" and "Routine occupations" as original, while the rest as "Other occupations". Here, by using the SAR in code format, we can make this much easier by using:

```{r}
sar_df <- sar_df %>% mutate(New_nssec = fct_other(nssec,   keep = c("6", "7", "8"),   other_level = "0" ))
```

Or by using `if_else` and `%in%` in R, we can achieve the same result. `%in%` is an operator used to test if elements of one vector are present in another. It returns `TRUE` for elements found and `FALSE` otherwise.

```{r}
sar_df <- sar_df %>% mutate(New_nssec = if_else(!nssec %in% c(6,7,8), "0" ,nssec))
```

Use "Other occupations" (code: 0) as the reference category by `relevel(as.factor())` and then create the regression model: `glm(New_work_distance~sex + New_nssec, data = sar_df, family= "binomial")`. Can you now run the model by yourself? Find the answer at the end of the practical.

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**Q4**. **If we want to explore whether people as independent employers show lower probability of commuting longer than 60km compared with other occupations, how will we prepare the input independent variables and what will be the specified regression model?**
:::

### Prediction using fitted regression model

Relating to this week's lecture, we can write down the fitted model. **Note** that the numbers below are the *odds ratios* returned by `exp(coef(m.glm))`, not the log-odds coefficients shown by `summary(m.glm)`, so they multiply together rather than add up:

Odds of long-distance commuting = 0.188 × (0.693 if female) × (0.679 if nssec3) × (0.357 if nssec4) × (3.409 if nssec5) × (0.249 if nssec6) × (0.237 if nssec7) × (0.226 if nssec8). We don't include nssec1, as the previous result shows it is not statistically significant, so we won't use the model to predict individuals in the nssec=1 occupation group.

By using R, you can create the object you would like to predict. Here we create three people; see whether you can interpret their gender and socio-economic classification.

```{r}
objs <- data.frame(sex=c("1","2","1"),nssec=c("7","3","5"))
```

Then we can predict by using our model `m.glm`:

```{r}
predict(m.glm, objs,type = "response")
```

So let us look at these three people. For the first, a male classified as Semi-routine occupation in NSSEC, the probability that he travels over 60km to work is only 4.26%. For the second, a female in a Lower managerial and professional occupation, the probability of long-distance commuting is 8.11%. Now work out the prediction for our last person. Remember: the model fit results show that the model is not very robust, so the predictions may not be very solid either.

## **Extension activities**

The extension activities are designed to help you prepare for the assignment. For this week, try whether you can:

-   Select a regression strategy and explain why a linear or logistic model is appropriate

-   Perform one or a series of regression models, including different combinations of your chosen independent variables to explain and/or predict your dependent variable

## **Answers for Qs**

**Answer for Q1**

For Q1, we have already had the codes for `work_distance` and `nssec`, so the only missing variable is `sex`:

To summarise the frequency of each category, we use the code:

```{r}
table(sar_label$sex)
```

To create bar charts for the distribution and composition of the gender variable in the SAR sample, we can run:

```{r}
ggplot(sar_label, aes(x = sex)) +
  geom_bar(fill="gold",width=0.5) +
  labs(title = "Histogram of Gender in SAR sample", x = "gender", y = "Count")+#set text info
  theme_classic()#choose theme type, try theme_bw(), theme_minimal() see differences
```

**Answer for Q2**

Similarly, the descriptive analysis for the newly created variable `New_work_distance` in `sar_df` should include:

```{r}
table(sar_df$New_work_distance)
```

Here we see the categories of `New_work_distance` are in numeric type 0 or 1 (as we used code `sar_df <- sar_df %>% mutate( New_work_distance = if_else(work_distance >6, 1,0))` to do so), therefore in the histogram code, we use as.factor() to convert it as factor type.

```{r}
ggplot(sar_df, aes(x = as.factor(New_work_distance))) +
  geom_bar(fill="forestgreen",width=0.5) +
  labs(title = "Histogram of New_work_distance in SAR sample", x = "New_work_distance", y = "Count")+#set text info
  theme_classic()#choose theme type, try theme_bw(), theme_minimal() see differences
```

Now we can use the frequency and the bar chart to report that according to our definition of long-distance travelling to work (over 60 km), there are 3,072 individuals who are long-distance commuters in the SAR sample, which makes up 6.14% (3072/50000) of the full sample.

**Answer for the model in Q4**

In Q4, we want to explore whether people whose occupation is independent employer are associated with a lower probability of commuting over long distances compared to people in other occupations. So we set nssec = 5 (Small employers and own account workers) as the baseline category.

```{r}
table(sar_df$New_nssec)
```

Then we set the reference categories: `sex` as `1` (male) and `nssec` as `5`, which is 'Small employers and own account workers':

```{r}
sar_df$sex <- relevel(as.factor(sar_df$sex),ref="1")
sar_df$nssec <- relevel(as.factor(sar_df$nssec),ref="5")
```

Now, we build the logistic regression model and check out the outcomes:

```{r}
model_new = glm(New_work_distance~sex + nssec, data = sar_df, family= "binomial")

summary(model_new)
```

For the model interpretation, we need:

```{r}
# odds ratios
exp(coef(model_new)) 

```

```{r}

# confidence intervals
exp(confint(model_new, level = 0.95)) 

```

And don't forget to measure the model fit:

```{r}

# model fit
pR2(model_new) %>% round(4) %>% tidy()
```

::: {style="background-color: #FFFBCC; padding: 10px; border-radius: 5px; border: 1px solid #E1C948;"}
**Q5.** **Now what conclusions can you draw when comparing the middle NSSEC groups to the higher groups?**
:::
